---
title: "Reproducible Research - Week2 "
author: "Pranab Niyogi"
date: "December 13, 2016"
output: html_document
---

### Setup directory, load libraries
```{r setup, include=FALSE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
setwd("C:/datascience/R/ReproducibleResearch/Week2")
```

### 1. Loading and preprocessing the data


```{r, echo=TRUE}

dfActivity <- read.csv("activity.csv");
dfActivity <- dfActivity[!is.na(dfActivity$steps),] 
 
```
 
### 2. What is mean total number of steps taken per day?

####  2.1 Show the Histogram


```{r, echo=TRUE}
dfActivitySum <- aggregate(dfActivity$steps, by=list(dfActivity$date),FUN=sum)
names(dfActivitySum) <- c("Date","Steps")
hist(dfActivitySum$Steps,breaks=10, main="Number of Steps", xlab="Steps per day")
```

####  2.2 Calculate and report the mean and median of the total number of steps taken per day

```{r, echo=TRUE}

dfActivitySumByDate <- aggregate(dfActivity$steps,by=list(dfActivity$date),FUN=sum)
names(dfActivitySumByDate) <- c("Date","steps")
# Mean
mean(dfActivitySumByDate$steps,na.rm=TRUE)
# Median
median(dfActivitySumByDate$steps,na.rm=TRUE)
```
### 3. What is the average daily activity pattern?

#### 3.1 Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
 
```{r, echo=TRUE}
dfActivityMeanByInterval <- aggregate(dfActivity$steps,by=list(dfActivity$interval),FUN=mean)
names(dfActivityMeanByInterval) <- c("Interval","Steps" )
 plot(dfActivityMeanByInterval$Interval,dfActivityMeanByInterval$Steps, type='l',xlab="Steps Interval",ylab="Average Steps")
```

#### 3.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
 
```{r, echo=TRUE}
dfActivityMeanByInterval[dfActivityMeanByInterval$Steps==max(dfActivityMeanByInterval$Steps),]
```
### 4. Imputing missing values

#### 4.1 Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
dfActivity <- read.csv("activity.csv");
nrow(dfActivity[is.na(dfActivity$steps),]) 
```
#### 4.2 Devise a strategy for filling in all of the missing values in the dataset. 
        Used average of 5 min interval accross dates to fillup the missing values

#### 4.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo=TRUE}
dfActivity <- read.csv("activity.csv");
dfActivity1 <- dfActivity[!is.na(dfActivity$steps),] 

avgperday <- aggregate(dfActivity1$steps,by=list(factor(as.character(dfActivity1$interval))), FUN=mean)
  for ( i in 1:nrow(dfActivity) )
  {
    if (is.na(dfActivity[i,1]))
    { 
      
      dfActivity[i,1] <- avgperday[which(dfActivity[i,3]==avgperday$Group.1),2] 
    }  
    
  }

```


#### 4.4 Make a histogram of the total number of steps taken each day

 
```{r, echo=TRUE}
dfActivitySumByDate <- aggregate(dfActivity$steps,by=list(dfActivity$date),FUN=sum)
names(dfActivitySumByDate) <- c("Date","steps")
hist(dfActivitySumByDate$steps,breaks=10, main="Number of Steps", xlab="Steps per day")

```

#### 4.4.1 Calculate and report the mean and median total number of steps taken per day.Do these values differ from the estimates from the first part of the assignment? 
```{r, echo=TRUE}
# Mean
mean(dfActivitySumByDate$steps,na.rm=TRUE)
# Median
median(dfActivitySumByDate$steps,na.rm=TRUE)
```
#### 4.4.2 What is the impact of imputing missing data on the estimates of the total daily number of steps?
#### Estimating missing data Did not skewed the statistics.


### 5.0 Are there differences in activity patterns between weekdays and weekends?


#### 5.1 Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
dfActivity$day <- "weekday"
for (i in 1:nrow(dfActivity) ) 
{
  if ( weekdays(as.Date(dfActivity[i,2]),abb=T) %in% c("Sat","Sun") )
  {
   dfActivity[i,4] <- "weekend"
  }
}
```


#### 5.2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r, echo=TRUE}
library(lattice)
dfActivityByDOW <- aggregate(dfActivity$steps,by=list(dfActivity$interval,dfActivity$day), FUN="mean")
names(dfActivityByDOW) <- c("interval","day","steps")
xyplot(steps ~ interval | day, data=dfActivityByDOW, type="l", layout=c(1,2), xlab="Interval",ylab="Steps")
```
